---
title: "Intro to Raster Data"
format: html
---

## Data

The data is described here: https://datacarpentry.org/geospatial-workshop/#data.

```{r message=FALSE}
library(terra)
library(ggplot2)
library(dplyr)
library(jsonlite)
```

https://www.neonscience.org/about

The [**raster** package]() was the package of choice but now both [**terra**]() and [**stars**]() seem to have
supplanted it. The same person (Robert Hijmans) who wrote **raster** is also the main dev for **terra**. It seems that **terra** is somewhat more efficient (speed or space?). The primary data classes in **raster** are the `RasterLayer`, `RasterStack` (a collection of `RasterLayer` objects under the hood) and `RasterBrick` (true multi-layer object but can only reference a single file). In **terra**, the `SpatRaster` is the class that handles single and multi-layer raster objects. There are functions for translating between the the two packages.

There's a very good comparison of **terra** and **stars** in [the Raster Data chapter of Geocomputation with R](https://r.geocompx.org/spatial-class.html#r-packages-for-working-with-raster-data). The **terra** package is somewhat simpler and focuses on regular rectangular rasters of one or more layers. The **stars** package can handle [more complex raster shapes](https://r-spatial.github.io/stars/) as well as data cubes (several bands, a time dimension, x and y coordinates, many attributes). The Python package **xarray** seems similar to **stars** and **terra**. In **stars**,
the vector data classes are the same as in the **sf** package whereas **terra** uses its own `SpatVector` class.



## View raster file attributes

The raster files are GeoTIFF files. We can view overall file info via
the [GDALInfo tool](https://gdal.org/programs/gdalinfo.html). It's wrapped by the `terra::describe()` function.

```{r}
HARV_dsmCrop_file <- "data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif"
terra::describe(HARV_dsmCrop_file)
```

We can store the information in a character vector using the utils function
`capture.output()`.

```{r capture_gdal_info}
HARV_dsmCrop_info <- capture.output(
  describe("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")
)
```

## Open a raster file in R

To load a GeoTIFF raster file into a SpatRaster file, we use the `terra::rast()` function.

```{r load_raster}
DSM_HARV <-
  rast("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")

DSM_HARV
```

The `SpatRaster` class is a way of representing multi-layer raster file. This
file only has one layer. A `SpatRaster` feels similar to a Python xarray `DataArray`.

```{r}
summary(DSM_HARV)
```

Use `values` to force all cells in the raster to be considered by `summary()`.

```{r}
summary(values(DSM_HARV))
```

To plot a `SpatRaster` using ggplot2, we need to create a dataframe. The
`xy = TRUE` specifies that we want the coordinates included.

```{r}
DSM_HARV_df <- terra::as.data.frame(DSM_HARV, xy = TRUE)
str(DSM_HARV_df)
```

The coordinate values are eastings (x) and northings (y) for our [UTM zone 18N projection](https://www.usgs.gov/faqs/what-does-term-utm-mean-utm-better-or-more-accurate-latitudelongitude).

Now we can use ggplot2 to plot our raster data. A few things to note:

- the `fill` aesthetic are the raster values (elevation values in this case),
- for the color pallete we are using `scale_fill_viridis_c()` which is color blindness friendly,
- the `coord_quickmap()` is a quick approximate projection that works well with small areas that aren't too close to the poles.

```{r raster_plot}
ggplot() +
    geom_raster(data = DSM_HARV_df , aes(x = x, y = y, fill = HARV_dsmCrop)) +
    scale_fill_viridis_c() +
    coord_quickmap()
```
There is a built in `plot` function in the terra package that can be used
directly on the SpatRaster object.

```{r}
plot(DSM_HARV)
```
While we can tell from the raster file metadata output that the easting and
northing values are in meters. What are the units for the elevation values? This
information should be (?) part of the coordinate reference system. We can get
at the details through the `terra::crs()` function.

```{r}
crs(DSM_HARV, proj = TRUE)
```

We see that the units are "m", which is meters. Of course, this applies to 
the x and y coordinates, but the raster values themselves aren't necessarily
related to the CRS.

There are other ways to plot `SpatRaster` objects, including:

- **terra** has a `plotRGB` function to easily plot RGB (3-band) raster files,
- the **tmap** package which can create both static and interactive plots
- the **rasterVis** package to do faceted plotting (useful to time-lapse plots)



## Raster bands

This raster has only one band. Other rasters such as the 'visual' item form
Sentinel-2 contain three bands - red, green and blue (already scaled to 0-255).

```{r read_sentinal_rgb}
rgb_sentinal_example_file <- 'data/iargo_springs.tif'
terra::describe(rgb_sentinal_example_file)
```
Now let's try a Landsat generated file.


```{r plot_rgb}

rgb_landsat_example_file <- 'data/NEON-DS-Landsat-NDVI/SJER/2011/RGB/014_SJER_landRGB.tif'

# Read into SpatRaster
rgb_landsat_example <- rast(rgb_landsat_example_file)

# Plot it
terra::plotRGB(rgb_landsat_example, r=1, g=2, b=3, scale=255)
```
Hmmm, seems a little strange.



## Dealing with missing data

There is of a `NoDataValue` assigned to raster data to deal with pixels for which
there is no available data. This can be caused by how the data was collected (e.g. the satellite sweep)
or other data capture issues. Let's see if this raster file has a such a no data value.

We can use the metadata we saw earlier. The `capture.output()` function doesn't really
capture the output of `describe()` very well as we don't even get a named character vector.
There's got to be a better way to get the output of GDALInfo().

```{r}
HARV_dsmCrop_info[57:60]
```

```{r}
gdalinfo_output <- sf::gdal_utils("info", HARV_dsmCrop_file)
```

```{r}
class(gdalinfo_output)
is.vector(gdalinfo_output)
names(gdalinfo_output)
```

What about json? Newer versions of GDAL write write json output for the `gdalinfo()` function.
https://themockup.blog/posts/2020-05-22-parsing-json-in-r-with-jsonlite/

```{r}
gdalinfo_output_json <- sf::gdal_utils("info", HARV_dsmCrop_file, options = c("-json"), quiet = TRUE)

gdal_info <- jsonlite::fromJSON(gdalinfo_output_json)

gdal_info$coordinateSystem$wkt
```

Let's write out the JSON.

```{r}
fileConn<-file("data/gdalinfo_output.json")
writeLines(gdalinfo_output_json, fileConn)
close(fileConn)
```

We can also view it with a built in JSON viewer in R Studio. Well, we
don't get anything useful.

```{r}
View(gdalinfo_output_json)
```


## Distribution of elevation values

We can use ggplot2 but need to use the dataframe version of our data.

```{r}
ggplot(terra::as.data.frame(DSM_HARV)) + geom_histogram(aes(x = HARV_dsmCrop), bins=40)
```

## Exercise

Challenge: Explore Raster Metadata

Use describe() to determine the following about the NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_DSMhill.tif file:

    Does this file have the same CRS as DSM_HARV?
    What is the NoDataValue?
    What is resolution of the raster data?
    How large would a 5x5 pixel area be on the Earth’s surface?
    Is the file a multi- or single-band raster?

Notice: this file is a hillshade. We will learn about hillshades in the [Working withMulti-band Rasters in R episode.


```{r}
HARV_dsmHill_file <- "data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_DSMhill.tif"
HARV_dsmHill <- rast(HARV_dsmHill_file)
terra::describe(HARV_dsmHill_file)
```

Does this hillshade file use the same CRS as the elevation file?

```{r}
crs(HARV_dsmHill) == crs(DSM_HARV) 
```

What is the `NoDataValue`?

> The output of `describe` (i.e. `gdalinfo()`) shows it to be -9999. 

What is the resolution?

> Pixel Size = (1.000000000000000,-1.000000000000000)
>
> My guess is that this is from the perspective of the upper left corner of
a pixel. The units are meters. So, the first 1.0 is the length of the pixel
along the edge connecting upper left and upper right. The -1.0 value is
the length of the edge from upper left to lower left (hence, the negative sign).

Area of 5x5 pixels?

> 25m^2

Single or multi-band raster?

> Single (see gdalinfo() output). We can also do check it via **terra** functions.

```{r}
dim(DSM_HARV)
nlyr(DSM_HARV)
```

# Ch 2 - Plotting raster data

Objectives are:

- Build customized plots for a single band raster using the ggplot2 package,
- Layer a raster dataset on top of a hillshade to create an elegant basemap.

Let's start by binning the elevation field. Then we can use color to show the
elevation "class".

```{r}
DSM_HARV_df <- DSM_HARV_df %>%
                mutate(fct_elevation = cut(HARV_dsmCrop, breaks = 3))

ggplot() +
    geom_bar(data = DSM_HARV_df, aes(x = fct_elevation))
```
The levels for the new `fct_elevation` field are:

```{r}
unique(DSM_HARV_df$fct_elevation)
```

```{r}
DSM_HARV_df %>% 
  group_by(fct_elevation) %>% 
  count()
```

Let's make nicer breakpoints.

```{r}
custom_bins <- c(300, 350, 400, 450)

DSM_HARV_df <- DSM_HARV_df %>%
  mutate(fct_elevation_2 = cut(HARV_dsmCrop, breaks = custom_bins))

unique(DSM_HARV_df$fct_elevation_2)

DSM_HARV_df %>% 
  group_by(fct_elevation_2) %>% 
  count()
```

Now we can plot it using `geom_raster`, which is built in to **ggplot2**. There
are related functions for plotting rectangles and tiles.

```{r}
ggplot() +
  geom_raster(data = DSM_HARV_df , aes(x = x, y = y, fill = fct_elevation_2)) + 
  coord_quickmap()
```

We can customize the colors for terrain plots using the `terrain.colors` function along
with `scale_fill_manual` from **ggplot2**.

```{r}
terrain.colors(3)
```

```{r}
ggplot() +
 geom_raster(data = DSM_HARV_df , aes(x = x, y = y,
                                      fill = fct_elevation_2)) + 
    scale_fill_manual(values = terrain.colors(3)) + 
    coord_quickmap()
```
Make things more generic and a little nicer with axis labels. You can
suppress axis labels with `theme(axis.title = element_blank())`.

```{r}
my_col <- terrain.colors(3)

ggplot() +
 geom_raster(data = DSM_HARV_df , aes(x = x, y = y,
                                      fill = fct_elevation_2)) + 
    scale_fill_manual(values = my_col, name = "Elevation") + 
    coord_quickmap() + xlab("Easting") + ylab("Northing")
```

## Layering rasters

> We can layer a raster on top of a hillshade raster for the same area, and use a transparency factor to create a 3-dimensional shaded effect. A hillshade is a raster that maps the shadows and texture that you would see from above when viewing terrain. We will add a custom color, making the plot grey.

Read in a hillshade raster.

```{r}
DSM_hill_HARV <-
  rast("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_DSMhill.tif")

DSM_hill_HARV
```

Convert to dataframe for plotting.

```{r}
DSM_hill_HARV_df <- as.data.frame(DSM_hill_HARV, xy = TRUE) 

str(DSM_hill_HARV_df)
```

For the raster plot of the hillshade, we'll use the values as the `alpha` argument (transparency)
in our aesthetic mapping. Since the raw `HARV_DSMhill` values aren't all valid, or desirable, transparency
values, we'll rescale the values to all be between 0.15 and 0.65 (the default is 0.1 - 0.9). 

```{r}
ggplot() +
  geom_raster(data = DSM_hill_HARV_df,
              aes(x = x, y = y, alpha = HARV_DSMhill)) + 
  scale_alpha(range =  c(0.15, 0.65), guide = "none") + 
  coord_quickmap()
```
Now combine them in a layered plot.

```{r}
ggplot() +
  geom_raster(data = DSM_HARV_df , 
              aes(x = x, y = y, 
                  fill = HARV_dsmCrop)) + 
  geom_raster(data = DSM_hill_HARV_df, 
              aes(x = x, y = y, 
                  alpha = HARV_DSMhill)) +  
  scale_fill_viridis_c() +  
  scale_alpha(range = c(0.15, 0.65), guide = "none") +  
  ggtitle("Elevation with hillshade") +
  coord_quickmap()
```